Leaf mass per area, not total leaf area, drives differences in above-ground biomass distribution among woody plant functional types

### 2016-04-22 manuscript.Rnw
Move Fig S2 into panel of Fig 3

Using variance partitioning, we quantified the contribution of plant height, plant functional type, and climate to the overall variation in $\mfms$ and for all three component variables. The amount of variance explained by quantitative climate variables (MAP and MAT) was assessed with GAMs, with variables sequentially added to the model and the explained variance ($R^2$) calculated. We also compared this approach to another method using linear mixed-effects models instead of GAMs \citep[calculating the $R^2$ for linear mixed-effects models for the fixed effects only following][]{nakagawa2013}, as well as third method called ``independent effects analysis''. The three methods generally agreed on the ranking of variable importance (see Fig. S5 and S6).

@@

As expected, both whole-plant leaf mass ($\mf$) and above-ground woody biomass ($\ms$) showed a strong dependence on plant height (Fig. 2). For both variables the relationship was non-linear on a log-log scale, demonstrating departure from simple allometric relationships. It is also evident that $\ms$ increases much quicker with plant height than $\mf$. As a result, the ratio $\mfms$ decreased with plant height (Fig. 3a), from an average of \Sexpr{mfmssmall} for very small saplings (\textless{} 0.1m) to an average of \Sexpr{mfmslarge} for large trees (\textgreater{} 30m). Similarly, $\afms$ decreased with plant height (Fig. S2), from an average of \Sexpr{afmssmall} for very small saplings to an average of \Sexpr{afmslarge} for large trees.

@@

We found large differences between the three major PFTs in terms of above-ground biomass distribution, as quantified by the ratio $\mfms$ (Fig. 3a), but not in the ratio $\afms$ (Fig. S2). Differences in $\mfms$ were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater{} evergreen angiosperm \textgreater{} deciduous angiosperm. Next we calculated $\mfms$ at a common plant height of 3m for each of the PFTs (the mean height across all data, \Sexpr{format(10^mean(dataset$lh.t, na.rm=TRUE), nsmall=1, digits=2)}m). These three size-standardised values of $\mfms$ were highly significantly different between PFTs (p \textless{} 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig. 3b). As a result, the amount of leaf area per unit above-ground woody biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater{} 0.05, see Fig. 3c), because the differences in $\mfms$ were exactly compensated by differences in LMA.

These patterns in biomass distribution reflected differences in the distribution of components of eq. \ref{eq:mfmspart}-\ref{eq:afmspart}. In particular, LMA was found to differ substantially among PFTs, whereas $\afas$ did not (Fig. 4), which is consistent with the large differences between PFTs in LMA shown in Fig. 3b,c.  When PFT was added to the statistical model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mfaf$ but not $\afas$  (Table \ref{table:varparttable}), demonstrating that PFT differences entered largely in leaf mass-based variables.

@@

Climate variables consistently added little explanatory power when added to a statistical model that included PFT and height (the adjusted R$^2$ frequently decreased, see Table \ref{table:varparttable}). The only exception was LMA, for which a model that included MAT explained 53\% of the variation, compared to a model with height and PFT (45\%). Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citet{reich2014}, we also analysed climate effects with methods similar to that study. When we tested for MAT effects on $\mfms$ when only considering gymnosperms and angiosperms (thus grouping deciduous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig. S5) for angiosperms for most size classes (but never for gymnosperms). We found, however, that this was largely due to the fact that at higher MAT, an increasing proportion of angiosperms are evergreen (Fig. 1), which have higher $\mfms$. When analysed separately for deciduous and evergreen angiosperms, few and weaker significant relationships were found (Fig. S6).

@@

We found that $\mfms$ steeply declined with plant height. It is well known that it is necessary to correct for plant size when comparing biomass distribution among groups \citep{mcconnaughay1999}. We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship. This was useful because both $\mf$ and $\ms$ showed very non-linear patterns with plant height (Fig. 2), or among biomass pools (Fig. S4), even on a logarithmic scale. This is consistent with recent results on an independently compiled large database of plant allometry \citep{poorter2015}, and further calls into question the generality of the often-assumed log-linear relationships between plant biomass pools and size metrics.

We have confirmed the hypothesis that the amount of leaf area supported per plant is less variable between PFTs than the amount of leaf mass. \citet{poorter2015} hypothesised such a result might arise, because forests would converge to a similar leaf area index (LAI). Although we found remarkably similar leaf area per unit above-ground woody biomass ($\mfms$) between the three PFTs studied (Fig. S2), this does not in itself imply that LAI is equally similar. Indeed, in a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI.  Moreover, we would in fact expect LAI to change with LMA, not because of changes in the amount of leaf area maintained by individual plants, but because LMA correlates with the shade-tolerance of seedlings \citep{poorter2006, falster2011}, and thereby the LAI of the community \citep{falster2011}. The ultimate reason LMA might affect shade tolerance is because is via its correlation with leaf lifespan \citep{wright2004}. For whole stands, we can therefore expect total stand leaf mass to be correlated with LMA, both because of greater leaf mass per plant \citep[e.g.]{reich1992}, and the greater density of plants arising from increased leaf lifespan. Previous interpretations of differences in $\mfms$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002, poorter2012}. We argue that there are in fact two effects occurring simultaneously: differences in LMA cause differences in the $\mfms$ of individual plants, while differences in LL (and thereby LMA) cause differences in the LAI of whole stands.

After accounting for PFT and plant height, we found very weak and inconsistent effects of climate on all variables studied, with some exception for LMA (Table \ref{table:varparttable}). Climate was represented as either mean annual temperature or an aridity index (PET/MAP), however, results were similarly inconclusive when studying other climate variables such as composite drought indices, growing season length, or mean annual precipitation (results not shown). This result appears to contradict \citet{reich2014} who reported consistent increases in foliage biomass ($\mf$) with MAT, when corrected for above-ground woody biomass ($\ms$). We were able to partly reconcile our findings with those of \citet{reich2014}: when we categorised our species as only angiosperms vs gymnosperms (thus grouping deciduous and evergreens, as did Reich \emph{et al.}), $\mfms$ increased consistently with MAT for most size classes within angiosperms (Fig. S5), in line with the findings of \citet{reich2014}; on the other hand we found no comparable relationship for gymnosperms. Moreover, we found that in our dataset, the significant relationships observed for angiosperms as a whole, could be explained via differences between the LMA of deciduous and evergreen species: evergreen angiosperms occurred both at higher MAT (Fig. 1), and had a higher LMA (Fig. 2), resulting in a correlation between $\mfms$ and MAT. We also re-analysed the  data in Reich et al. 2014 (not shown) to determine whether this shift in PFT with climate could also explain their finding of an increase in $\mfms$ with MAT, but found the pattern still held within gymnosperms and deciduous angiosperms.

@@

\textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
\textbf{Fig. S2} Leaf area ratio ($\afms$) by plant height and PFT.\\
\textbf{Fig. S3} Comparison of three methods for variance partitioning of the four studied variables. \\
\textbf{Fig. S4} Whole-plant leaf mass ($\mf$) and leaf area ($\af$) as a function of total aboveground biomass. \\
\textbf{Fig. S5} Relationships between leaf - stem biomass distribution ($\mfms$) and mean annual temperature (MAT) (\degree C). \\
\textbf{Fig. S6} As Fig. S5 but with angiosperms split up into evergreen and deciduous. \\

>>>>

Using variance partitioning, we quantified the contribution of plant height, plant functional type, and climate to the overall variation in $\mfms$ and for all three component variables. The amount of variance explained by quantitative climate variables (MAP and MAT) was assessed with GAMs, with variables sequentially added to the model and the explained variance ($R^2$) calculated. We also compared this approach to another method using linear mixed-effects models instead of GAMs \citep[calculating the $R^2$ for linear mixed-effects models for the fixed effects only following][]{nakagawa2013}, as well as third method called ``independent effects analysis''. The three methods generally agreed on the ranking of variable importance (see Fig. S4 and S5).

@@

As expected, both whole-plant leaf mass ($\mf$) and above-ground woody biomass ($\ms$) showed a strong dependence on plant height (Fig. 2). For both variables the relationship was non-linear on a log-log scale, demonstrating departure from simple allometric relationships. It is also evident that $\ms$ increases much quicker with plant height than $\mf$. As a result, the ratio $\mfms$ decreased with plant height (Fig. 3a), from an average of \Sexpr{mfmssmall} for very small saplings (\textless{} 0.1m) to an average of \Sexpr{mfmslarge} for large trees (\textgreater{} 30m). Similarly, $\afms$ decreased with plant height (Fig. 3c), from an average of \Sexpr{afmssmall} for very small saplings to an average of \Sexpr{afmslarge} for large trees.

@@

We found large differences between the three major PFTs in terms of above-ground biomass distribution, as quantified by the ratio $\mfms$ (Fig. 3a), but not in the ratio $\afms$ (Fig. 3c). Differences in $\mfms$ were fairly consistent along the range in plant height, with a ranking of evergreen gymnosperm \textgreater{} evergreen angiosperm \textgreater{} deciduous angiosperm. Next we calculated $\mfms$ at a common plant height of 3m for each of the PFTs (the mean height across all data, \Sexpr{format(10^mean(dataset$lh.t, na.rm=TRUE), nsmall=1, digits=2)}m). These three size-standardised values of $\mfms$ were highly significantly different between PFTs (p \textless{} 0.001), and were approximately proportional to the average LMA ($\mfaf$) across the PFTs (Fig. 3b). As a result, the amount of leaf area per unit above-ground woody biomass ($\afms$) did not differ between PFTs (all pair-wise comparisons p \textgreater{} 0.05, see Fig. 3d), because the differences in $\mfms$ were exactly compensated by differences in LMA.

These patterns in biomass distribution reflected differences in the distribution of components of eq. \ref{eq:mfmspart}-\ref{eq:afmspart}. In particular, LMA was found to differ substantially among PFTs, whereas $\afas$ did not (Fig. 4), which is consistent with the large differences between PFTs in LMA shown in Fig. 3b.  When PFT was added to the statistical model (as an intercept term as well as interaction with the smooth plant height term), the $R^2$ increased substantially for $\mfaf$ but not $\afas$  (Table \ref{table:varparttable}), demonstrating that PFT differences entered largely in leaf mass-based variables.

@@

Climate variables consistently added little explanatory power when added to a statistical model that included PFT and height (the adjusted R$^2$ frequently decreased, see Table \ref{table:varparttable}). The only exception was LMA, for which a model that included MAT explained 53\% of the variation, compared to a model with height and PFT (45\%). Since the lack of climate effects on biomass distribution is in contrast with recent finding of \citet{reich2014}, we also analysed climate effects with methods similar to that study. When we tested for MAT effects on $\mfms$ when only considering gymnosperms and angiosperms (thus grouping deciduous and evergreen species), we found a significant and consistent positive relationship with MAT (Fig. S4) for angiosperms for most size classes (but never for gymnosperms). We found, however, that this was largely due to the fact that at higher MAT, an increasing proportion of angiosperms are evergreen (Fig. 1), which have higher $\mfms$. When analysed separately for deciduous and evergreen angiosperms, few and weaker significant relationships were found (Fig. S5).

@@

We found that $\mfms$ steeply declined with plant height. It is well known that it is necessary to correct for plant size when comparing biomass distribution among groups \citep{mcconnaughay1999}. We used a semi-parametric approach to account for plant size, which has the advantage that it does not require an \emph{a priori} assumption on the functional relationship. This was useful because both $\mf$ and $\ms$ showed very non-linear patterns with plant height (Fig. 2), or among biomass pools (Fig. S3), even on a logarithmic scale. This is consistent with recent results on an independently compiled large database of plant allometry \citep{poorter2015}, and further calls into question the generality of the often-assumed log-linear relationships between plant biomass pools and size metrics.

We have confirmed the hypothesis that the amount of leaf area supported per plant is less variable between PFTs than the amount of leaf mass. \citet{poorter2015} hypothesised such a result might arise, because forests would converge to a similar leaf area index (LAI). Although we found remarkably similar leaf area per unit above-ground woody biomass ($\mfms$) between the three PFTs studied (Fig. 3c), this does not in itself imply that LAI is equally similar. Indeed, in a global compilation of LAI for undisturbed forest stands, \citet{iio2014} reported a much higher LAI for evergreen gymnosperms compared to evergreen angiosperms, with deciduous angiosperms having intermediate and more variable LAI.  Moreover, we would in fact expect LAI to change with LMA, not because of changes in the amount of leaf area maintained by individual plants, but because LMA correlates with the shade-tolerance of seedlings \citep{poorter2006, falster2011}, and thereby the LAI of the community \citep{falster2011}. The ultimate reason LMA might affect shade tolerance is because is via its correlation with leaf lifespan \citep{wright2004}. For whole stands, we can therefore expect total stand leaf mass to be correlated with LMA, both because of greater leaf mass per plant \citep[e.g.]{reich1992}, and the greater density of plants arising from increased leaf lifespan. Previous interpretations of differences in $\mfms$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002, poorter2012}. We argue that there are in fact two effects occurring simultaneously: differences in LMA cause differences in the $\mfms$ of individual plants, while differences in LL (and thereby LMA) cause differences in the LAI of whole stands.

After accounting for PFT and plant height, we found very weak and inconsistent effects of climate on all variables studied, with some exception for LMA (Table \ref{table:varparttable}). Climate was represented as either mean annual temperature or an aridity index (PET/MAP), however, results were similarly inconclusive when studying other climate variables such as composite drought indices, growing season length, or mean annual precipitation (results not shown). This result appears to contradict \citet{reich2014} who reported consistent increases in foliage biomass ($\mf$) with MAT, when corrected for above-ground woody biomass ($\ms$). We were able to partly reconcile our findings with those of \citet{reich2014}: when we categorised our species as only angiosperms vs gymnosperms (thus grouping deciduous and evergreens, as did Reich \emph{et al.}), $\mfms$ increased consistently with MAT for most size classes within angiosperms (Fig. S4), in line with the findings of \citet{reich2014}; on the other hand we found no comparable relationship for gymnosperms. Moreover, we found that in our dataset, the significant relationships observed for angiosperms as a whole, could be explained via differences between the LMA of deciduous and evergreen species: evergreen angiosperms occurred both at higher MAT (Fig. 1), and had a higher LMA (Fig. 2), resulting in a correlation between $\mfms$ and MAT. We also re-analysed the  data in Reich et al. 2014 (not shown) to determine whether this shift in PFT with climate could also explain their finding of an increase in $\mfms$ with MAT, but found the pattern still held within gymnosperms and deciduous angiosperms.

@@

\textbf{Fig. S1} Global coverage of the climate space by the dataset, labelled by vegetation type.\\
\textbf{Fig. S2} Comparison of three methods for variance partitioning of the four studied variables. \\
\textbf{Fig. S3} Whole-plant leaf mass ($\mf$) and leaf area ($\af$) as a function of total aboveground biomass. \\
\textbf{Fig. S4} Relationships between leaf - stem biomass distribution ($\mfms$) and mean annual temperature (MAT) (\degree C). \\
\textbf{Fig. S5} As Fig. S4 but with angiosperms split up into evergreen and deciduous. \\

### 2016-04-22 manuscript_suppinfo.Rnw
Restructure introduction & results to emphasise hypotheses
- as requested by referees, move decomposition equation into introduction
- use this as basis for outlining hypotheses
- add tests in relation to af/ms as major result
- remove stem mass index from results and figures

Also, suggest

- moving Fig 6 - > Fig 4
- embedding Fig S2 in Fig 3
Move Fig S2 into panel of Fig 3

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
    \caption{Leaf area ratio ($\afms$) by plant height (H) and PFT. Each symbol is an individual plant. Lines are generalized additive model fits using species/dataset as a random effect. Shaded areas are approximate 95\% confidence intervals.}
    \label{fig:figureS2}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
    \caption{Comparison of three methods for variance partitioning of the four studied variables (see legend) by total plant height, plant functional type and climate (MAP, MAT and interaction). \textbf{GAM}: variance component as estimated with a generalized additive model, as the reduction in $R^2$ when the independent variable (i.e. PFT, climate or height) was excluded from a model. Note that a GAM does not include interactions. \textbf{LMM}: linear mixed model with all interactions, and squared terms for the continuous variables, where the variance component was again estimated as the reduction in $R^2$ when the independent variable (and all its interactions) were removed from the model. $R^2$ was calculated following \citet{nakagawa2013} using the `MuMIn` R package (excluding random effects) \citep{barton2015}. Note that LMM and GAM give very similar ranking, and even absolute variance components, for nearly all variables. Both methods identified $\afas$ and LMA as the only variables with a practically significant dependence on climate (variance component \textgreater{} 0.1). \textbf{IEA}: independent effects analysis (aka hierarchical partitioning), using the algorithm by \citet{chevan1991}, as implemented in the `hier.part` package for R \citep{walsh2013}. Note that this method does not include interactions or squared terms, yet gave similar results to GAM and LMM (except for variables that explained little variation).}
    \label{fig:figureS4}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
    \caption{Whole-plant leaf mass ($\mf$) and leaf area ($\af$) as a function of total aboveground biomass. In the main text, we have chosen to use plant height as a metric of plant size, to avoid correlations between dependent and independent variables ($\mf$ appears in both variables in panel a). The overall differences between PFTs, and the finding that leaf mass is more variable between PFTs than leaf area, still hold when we use mass as the size covariate.}
    \label{fig:figureS5}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
    \caption{Relationships between leaf - stem biomass distribution ($\mfms$) and mean annual temperature (MAT) (\degree C). Similar to \citet{reich2014}, we divided the data into five bins of plant size (bins were chosen so that each bin had equal sample size), and used only evergreen vs. deciduous plant functional types. Lines are simple linear regression fits across averages by species within dataset, but symbols are individual plant observations. Solid lines are significant (p \textless{} 0.05), dashed lines are not (p \textgreater{} 0.05).}
    \label{fig:figreichclimate}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS6.pdf}
    \caption{As Fig S5 but with angiosperms split up into evergreen and deciduous. Note how climate effects have all but disappeared, except for evergreen angiosperms in two size class bins.}
    \label{fig:figreichclimate2}
\end{figure}

>>>>

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS2.pdf}
    \caption{Comparison of three methods for variance partitioning of the four studied variables (see legend) by total plant height, plant functional type and climate (MAP, MAT and interaction). \textbf{GAM}: variance component as estimated with a generalized additive model, as the reduction in $R^2$ when the independent variable (i.e. PFT, climate or height) was excluded from a model. Note that a GAM does not include interactions. \textbf{LMM}: linear mixed model with all interactions, and squared terms for the continuous variables, where the variance component was again estimated as the reduction in $R^2$ when the independent variable (and all its interactions) were removed from the model. $R^2$ was calculated following \citet{nakagawa2013} using the `MuMIn` R package (excluding random effects) \citep{barton2015}. Note that LMM and GAM give very similar ranking, and even absolute variance components, for nearly all variables. Both methods identified $\afas$ and LMA as the only variables with a practically significant dependence on climate (variance component \textgreater{} 0.1). \textbf{IEA}: independent effects analysis (aka hierarchical partitioning), using the algorithm by \citet{chevan1991}, as implemented in the `hier.part` package for R \citep{walsh2013}. Note that this method does not include interactions or squared terms, yet gave similar results to GAM and LMM (except for variables that explained little variation).}
    \label{fig:figureS4}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS3.pdf}
    \caption{Whole-plant leaf mass ($\mf$) and leaf area ($\af$) as a function of total aboveground biomass. In the main text, we have chosen to use plant height as a metric of plant size, to avoid correlations between dependent and independent variables ($\mf$ appears in both variables in panel a). The overall differences between PFTs, and the finding that leaf mass is more variable between PFTs than leaf area, still hold when we use mass as the size covariate.}
    \label{fig:figureS5}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS4.pdf}
    \caption{Relationships between leaf - stem biomass distribution ($\mfms$) and mean annual temperature (MAT) (\degree C). Similar to \citet{reich2014}, we divided the data into five bins of plant size (bins were chosen so that each bin had equal sample size), and used only evergreen vs. deciduous plant functional types. Lines are simple linear regression fits across averages by species within dataset, but symbols are individual plant observations. Solid lines are significant (p \textless{} 0.05), dashed lines are not (p \textgreater{} 0.05).}
    \label{fig:figreichclimate}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{FigureS5.pdf}
    \caption{As Fig. S4 but with angiosperms split up into evergreen and deciduous. Note how climate effects have all but disappeared, except for evergreen angiosperms in two size class bins.}
    \label{fig:figreichclimate2}
\end{figure}
